---
title: "Clean AUS Football Data"
author: "Kodie McNamara"
date: "`r paste0('Last Updated: ', format(Sys.time(), '%d %B, %Y'))`"
output: html_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../output",
      output_file = "Clean NFL Data.html"
    )
  })
---

```{r setup, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
```


## Read in Data
```{r warning=FALSE}
nfl_data = read_xlsx(path = "../data/raw/AUS Sports Betting NFL Betting Data.xlsx") %>% as.data.frame()
str(nfl_data)
```

## Data Types

We transform certain columns to ensure they are of the correct data type

Each of the three variables are transformed to be categorical `Yes` or `No` values

1. Overtime?
2. Playoff Game?
3. Neutral Venue?


```{r}
nfl_data = nfl_data %>% 
  dplyr::mutate(
    `Overtime?` = factor(ifelse(is.na(`Overtime?`),"No","Yes")),
    `Playoff Game?` = factor(ifelse(is.na(`Playoff Game?`),"No","Yes")),
    `Neutral Venue?` = factor(ifelse(is.na(`Neutral Venue?`),"No","Yes"))
  )
```


## Convert NFL Teams

Full team names is less-than-deal. While the names won't be used in prediction, a three-letter identifier is more ideal. In addition, we would like to parse out the location of the given team.

```{r}
unique(c(nfl_data$`Home Team`, nfl_data$`Away Team`))
```

First, we parse out location of the game, since we can later on tie in things like region, weather, etc. for relevant predictor variables
```{r}
nfl_data$`Game Location` = sapply(
  nfl_data$`Home Team`,
  function(x){
    if(x=="Washington Football Team"){
      return("Washington")
    }
    return(
      paste0(head(strsplit(x," ")[[1]],-1), collapse = " ")
    )
  })

unique(nfl_data$`Game Location`)
```


With 37 teams, we provide a mapping from each team to a three-letter abbreviation, where we take historical team names and map them to current teams.
```{r}
map_name_to_abb = c("Indianapolis Colts" = "IND", 
                    "Philadelphia Eagles" = "PHI", 
                    "Kansas City Chiefs" = "KC", 
                    "San Francisco 49ers" = "SF", 
                    "Arizona Cardinals" = "ARZ", 
                    "Seattle Seahawks" = "SEA", 
                    "Carolina Panthers" = "CAR", 
                    "Cleveland Browns" = "CLE", 
                    "Jacksonville Jaguars" = "JAX", 
                    "Miami Dolphins" = "MIA", 
                    "New York Jets" = "NYJ", 
                    "Tennessee Titans" = "TEN", 
                    "Washington Commanders" = "WAS", 
                    "Minnesota Vikings" = "MIN", 
                    "Dallas Cowboys" = "DAL",
                    "Detroit Lions" = "DET", 
                    "Los Angeles Chargers" = "LAC", 
                    "Pittsburgh Steelers" = "PIT", 
                    "Denver Broncos" = "DEN", 
                    "Atlanta Falcons" = "ATL", 
                    "Baltimore Ravens" = "BAL", 
                    "Buffalo Bills" = "BUF", 
                    "Houston Texans" = "HOU", 
                    "New England Patriots" = "NE", 
                    "New Orleans Saints" = "NO", 
                    "New York Giants" = "NYG", 
                    "Green Bay Packers" = "GB", 
                    "Los Angeles Rams" = "LAR", 
                    "Las Vegas Raiders" = "LV", 
                    "Chicago Bears"= "CHI", 
                    "Tampa Bay Buccaneers" = "TB", 
                    "Cincinnati Bengals" = "CIN",
                    "Washington Football Team" = "WAS", 
                    "Washington Redskins" = "WAS", 
                    "Oakland Raiders" = "LV", 
                    "San Diego Chargers" = "LAC", 
                    "St. Louis Rams" = "LAR")

nfl_data$`Home Team` = map_name_to_abb[nfl_data$`Home Team`]
nfl_data$`Away Team` = map_name_to_abb[nfl_data$`Away Team`]

unique(c(nfl_data$`Home Team`, nfl_data$`Away Team`))
```

## Feature Creation

### Covering the Spread & Over/Under

The first feature to create is understanding if the home/away team covered the spread
```{r}
nfl_data$`Home Covered?` = (nfl_data$`Home Score` - nfl_data$`Away Score`) > nfl_data$`Home Line Close`
nfl_data$`Away Covered?` = (nfl_data$`Away Score` - nfl_data$`Home Score`) > nfl_data$`Away Line Close`
```


Next, we determine if the total score was over or under
```{r}
nfl_data$`Total Score Outcome` = ifelse((nfl_data$`Home Score` + nfl_data$`Away Score`) > nfl_data$`Total Score Close`,
                                        "Over",
                                        ifelse((nfl_data$`Home Score` + nfl_data$`Away Score`) < nfl_data$`Total Score Close`,
                                        "Under", 
                                        "Push"))
```

### Convert Decimal Odds to American Odds

We utilize the following conversion:

Let $\Omega_A = \text{American Odds}$ and $\Omega_D = \text{Decimal Odds}$. Then, we know
$$
\Omega_A = 
\left\{
\begin{array}{ll}
       \frac{-100}{\Omega_D-1} &  \Omega_D < 2 \\
       100 \times(\Omega_D - 1) & \Omega_D \geq 2 \\
\end{array} 
\right.
$$

Using this formula, we convert our decimal odd variables
```{r}
convert_decimal_to_american = function(x){
  return(
    round(ifelse(x<2, -100/(x-1), 100*(x-1)))
  )
}

variables_to_convert_odds_to_us = c(
  "Home Odds Open", "Home Odds Min", "Home Odds Max", "Home Odds Close", 
  "Away Odds Open", "Away Odds Min", "Away Odds Max", "Away Odds Close", 
  "Home Line Odds Open", "Home Line Odds Min", "Home Line Odds Max", 
  "Home Line Odds Close", "Away Line Odds Open", "Away Line Odds Min", 
  "Away Line Odds Max", "Away Line Odds Close")

for(var in variables_to_convert_odds_to_us){
  nfl_data[[var]] = convert_decimal_to_american(nfl_data[[var]])
}

rm(convert_decimal_to_american, variables_to_convert_odds_to_us,var)
```


### Line and Odds Movement

Next, define how much the line moved to determine if the crowd has any influence on the outcome
```{r}
nfl_data$`Home Line Movement` = nfl_data$`Home Line Close` - nfl_data$`Home Line Open` 
nfl_data$`Away Line Movement` = nfl_data$`Away Line Close` - nfl_data$`Away Line Open` 
```

Define how much the odds moved to determine if the crowd has any influence on the outcome
```{r}
nfl_data$`Home Odds Movement` = nfl_data$`Home Odds Close` - nfl_data$`Home Odds Open` 
nfl_data$`Away Odds Movement` = nfl_data$`Away Odds Close` - nfl_data$`Away Odds Open` 
```

### Date Variables

First, let's add a year, week, and day variable, which will not be used as a predictor
```{r}
nfl_data$`Day` = weekdays(nfl_data$Date)
nfl_data$Week = format(nfl_data$Date, "%V")
nfl_data$Year = format(nfl_data$Date, "%Y")
```


Next, we format the weeks. In particular:  

- Push Monday games back to the previous week, so we can account for them all in the same week
```{r}
nfl_data$Week = ifelse(nfl_data$Day=="Monday", as.numeric(nfl_data$Week)-1,nfl_data$Week)
```

Afterward, we format the years. In particular:

- Anything after Jan. 1, but before the next season (typically the beginning of September) belongs to the previous season

```{r}
nfl_data$Year = ifelse(nfl_data$Week < 26, as.numeric(nfl_data$Year) - 1, nfl_data$Year)
```


And finally since the years are correctly formatted, we can now

- Normalize each year so the first week is week 1 and not the calendar week
```{r}
nfl_data2 = nfl_data %>%
  mutate(Week = as.numeric(Week)) %>% 
  group_by(Year) %>% 
  mutate(Week = ifelse(Week - min(Week[Week>26]) < 0,
                       Week - min(Week[Week>26]) + max(Week),
                       Week - min(Week[Week>26]))) %>% 
  ungroup() %>% 
  mutate(Week = Week + 1)
```



### Primetime
We know all games played on Monday and Thursday are primetime, so let's identify that.
```{r}
nfl_data$`Primtetime?` = ifelse(is.element(nfl_data$Day,c("Thursday","Monday")), "Yes", "No")
```









































